>

### 2024-11-26 docs/supplementary.qmd
Move the daily-variation figure to the supplementary #24

---
title: >
  A temporal network analysis of drug co-prescription around antidepressants and anxiolytics uses in the Netherlands 2018-2022

format:
  elsevier-pdf:
    keep-tex: true
    include-in-header:
      file: _header.tex
    journal:
      formatting: review
      cite-style: authoryear
  docx:
    pandoc-options: ["--mathml"]

funding: "The author(s) received no specific funding for this work."
keywords:
  - Graph Theory
  - Prescription Registry
  - Polypharmacy
  - Mental Health   
  - Public Health Monitoring

metadata-files:
  - _author.yml
  - _abstract.yml
  - _knitr-bib.yml

in:
  article: true
---

@@

# Introduction

::: {.content-hidden}

Mental disorders are a common comorbidity affecting nearly two billion people worldwide with a prevalence that remains relatively stable overtime [@Kessler2009; @Steel2014; @Richter2019]. Having a chronic mental disorder is one factor which may lead to polypharmacy, *viz.* the use of multiple medications in treating complex clinical conditions [@Halli2019]. Polypharmacy is a prominent issue in modern medicine with a prevalence of 37% [@Delara2022], mainly attributed to the presence of multiple long-term comorbidity, i.e. multimorbidity, in the population. In resolving multimorbidity, the general edict of the evidence-based medicine is to follow an established clinical guideline to treat a particular clinical condition. However, these guidelines are tailored for a single disease without considering multimorbidity, resulting in the use of one or more medications per condition, leading to polypharmacy regimens for one patient with multiple conditions [@Molokhia2017]. Albeit clinical improvements offered by polypharmacy, risks of drug interaction and adverse effects also become more apparent, engendering a public health issue. Hitherto, no indicator of polypharmacy has been universally standardized [@Sirois2016], where the number of prescription is merely relied on as a crude metric [@Molokhia2017; @Delara2022].

:::

The relational characteristics in health research variables have fueled the growing implementation of network analysis, making it an indispensable research framework in medicine and epidemiology [@Luke2007]. Past implementations of network analysis in public health mostly focused on transmission network, social network, and organizational network. Only recently, @Cavallo2012 demonstrated a novel framework to describe medication uses on a population level, i.e. drug prescription network (DPN), a term which later is coined by @Bazzoni2015. Network analysis and graph theory offer an approach to model the relationship among building blocks of a complex system. From a complex system standpoint, polypharmacy is a prescription of multiple drug combinations which can be modelled as a network [@Miglio2021]. DPN captures a complex and dynamic system that may vary overtime, which is highly relevant to drug prescribed for mental disorders.

In psychiatry, polypharmacy is defined as a patient with at least two concurrent psychiatric medications [@Shrivastava2013]. With a prevalence between 13-90%, psychiatric polypharmacy manifests in five categories, namely same-class, multi-class, adjunctive, augmentation, and total polypharmacy. Same-class polypharmacy is the use of multiple medications from the same class. Multi-class polypharmacy is the use of multiple medications from different classes indicated for the same symptom cluster. Adjunctive polypharmacy is the use of additional medications to treat side effects due to other medications. Augmentation polypharmacy is the use of full-dose and sub-dose medications from a different class for the same symptom cluster. Finally, total polypharmacy is the total count of all medications used by a patient, both psychopharmaca and other classes alike [@Shrivastava2013]. For policymakers, differences in defining polypharmacy pose a challenge to design a monitoring system [@Sirois2016].

Monitoring polypharmacy regimens requires a holistic approach since drug prescription highly depends on the physician's prescribing habits [@Shrivastava2013], patient-physician relationship, and pharmaceutical market, which all bear an interplay with the decision-making process [@Cavallo2012]. A holistic polypharmacy monitoring system needs to describe the importance of polypharmacy by incorporating both qualitative and quantitative measures [@Sirois2016]. Quantitatively attributing importance to each medication is challenging, accounting for polypharmacy as a complex and interlinked medication uses. To adequately address the complexity, network analysis is a viable framework for analyzing polypharmacy. Transforming prescription registry into a DPN has been previously investigated, where DPN-based approach represented the general trend in co-prescription [@Cavallo2012; @Bazzoni2015].

A DPN is constructed as a matrix which rows to columns represent pairs of medication being taken by a patient in a day. Since there is no limit for the number of rows and columns, a DPN can contain virtually every possible combination of polypharmacy regimens. A DPN is generalizable as a population aggregate by coalescing all single-patient DPN through matrix addition [@Cavallo2012]. Constructing a DPN allows examining co-prescription dynamics in the population as graph measures, which is also attributable to each medication. One of such measures is graph centrality, where we calculate scores representing the number of co-medications, number of connecting paths, average distance, and relative importance [@Askar2021]. The relative importance score is of particular interest to address concerns raised by @Sirois2016 and @Delara2022 regarding standardized quantitative indicator of polypharmacy. Relative importance, most commonly measured by eigenvector centrality, represents the connection to vital medications. A high eigenvector centrality in a DPN indicates more prescriptions with highly-prescribed medications. Following up @Cavallo2012 and @Bazzoni2015, we hypothesize that eigenvector centrality as a measure of relative importance is attributable to indicate polypharmacy. Our investigation showcases eigenvector centrality as a quantitative indicator of polypharmacy to promote a systematic public health monitoring system.

::: {.content-hidden}
> **"How do we implement a network analysis approach to quantitatively attribute polypharmacy?"**
:::

# Methods

## Source of data

{{< include exec-summary/_source-data.qmd >}}

## Data pre-processing

{{< include exec-summary/_preprocessing.qmd >}}

## Measuring relative importance

{{< include exec-summary/_measure-eigen.qmd >}}

## Data analysis

### Exploratory descriptive analysis

{{< include exec-summary/_descriptive.qmd >}}

### Exploring seasonality in the dataset

{{< include exec-summary/_seasonality.qmd >}}

### Decomposition with singular spectrum analysis

{{< include exec-summary/_spectral.qmd >}}

### Determining relative importance

Since the sum of all eigenvector centralities $c_i$ in equation [-@eq-eigen-centrality] is equal to one, a network with similarly connected nodes will have an eigenvector centrality $c_i$ = $\frac{1}{n}$ for each node, where $n$ is the number of nodes. The value of $c_i = \frac{1}{n}$ is the expected eigenvector centrality in a densely connected network where connection $\mathcal{V}_{i, j}$ from node $\mathcal{V}_i$ to node $\mathcal{V}_j$ has the same probability for every node. Per the central limit theorem, the value $\frac{1}{n}$ can be parameterized as a mean of a normal distribution, where all instances of eigenvector centrality will follow the distribution, $c_i \sim \mathcal{N}(\mu, \sigma)$. A one-sample Student's T-test was performed to assess how much each eigenvector centrality differs from the expected values. As such, eigenvector centrality greater than $\frac{1}{n}$ can be considered high, and the value below it can be considered low. The p-values was corrected for a type-I error using Bonferroni method.

# Results

tar_load(res_tbl_overview)

n_patient <- prettyNum(res_tbl_overview$total, big.mark = ",", scientific = FALSE)
hi_claim <- res_tbl_overview$overview |> dplyr::slice_max(mean_claim)
lo_claim <- res_tbl_overview$overview |> dplyr::slice_min(mean_claim)

hi_mean <- prettyNum(hi_claim$mean_claim, big.mark = ",")
lo_mean <- prettyNum(lo_claim$mean_claim, big.mark = ",")

@@

IADB recorded `r n_patient` patients having at least one prescription of antidepressants or anxiolytics within five years of data extraction. @tbl-overview-patient depicts the dynamics of the population from 2018 to 2022. The ratio of male to female did not vary by a large margin, and the average age steadily increased over the year. These findings imply that the cohort stays relatively stable overtime without any indication of sudden demographic changes.

```{r tbl-overview-patient}
#| tbl-cap: "Number of participating patients with at least one claim of antidepressants or anxiolytics from 2018 to 2022"

tar_read(res_tbl_patient) |>
  dplyr::mutate(
    "year"       = as.character(year),
    "Total"      = n_male + n_female,
    "male"       = sprintf("%s (%s)", prettyNum(n_male,   big.mark = ","), scales::percent(p_male)),
    "female"     = sprintf("%s (%s)", prettyNum(n_female, big.mark = ","), scales::percent(p_female)),
    "male_age"   = sprintf("%.2f (%.2f)", age_male, age_sd_male),
    "female_age" = sprintf("%.2f (%.2f)", age_female, age_sd_female)
  ) |>
  dplyr::select(year, male, female, Total, male_age, female_age) |>
  flextable::flextable() |>
  flextable::set_header_labels(
    year       = "Year",
    male       = "Male",
    female     = "Female",
    male_age   = "Male",
    female_age = "Female"
  ) |>
  flextable::add_header_row(
    values    = c("", "Number of Patients (%)", "Mean Age (SD)"),
    colwidths = c(1, 3, 2),
    top       = TRUE
  ) |>
  flextable::align(align = "center", part = "all") |>
  flextable::align(j = 1, align = "left", part = "all") |>
  #flextable::font(fontname = "Times", part = "all") |>
  flextable::fontsize(size = 7, part = "all") |>
  flextable::autofit(add_w = 0, add_h = 0) |>
  flextable::set_table_properties(opts_pdf = list("arraystretch" = 1, "tabcolsep" = 0))

```{r tbl-res-poly}
#| tbl-cap: Summary statistics of different psychiatric polypharmacy categories

tar_read(res_tbl_poly) |>
  dplyr::mutate(
    "year" = as.character(year),
    "mean_sd" = sprintf("%.2f [%.2f]", mean, sd),
    "median_iqr" = sprintf("%.2f (%.2f)", median, IQR)
  ) |>
  dplyr::select(type:medication, sum, mean_sd) |>
  tidyr::pivot_wider(names_from = type, values_from = c(sum, mean_sd)) |>
  dplyr::select(year, medication, sum_same, sum_multi, mean_sd_same, mean_sd_multi, dplyr::contains("total")) |>
  flextable::flextable() |>
  flextable::set_header_labels(
    year = "Year",
    medication = "Medication",
    sum_same = "Same-class",
    sum_multi = "Multi-class",
    mean_sd_same = "Same-class",
    mean_sd_multi = "Multi-class",
    sum_total = "N",
    mean_sd_total = "Mean [SD]"
  ) |>
  flextable::add_header_row(
    values    = c("", "Population Level (N)", "Per Person (Mean [SD])", "Total Polypharmacy"),
    colwidths = c(2, 2, 2, 2),
    top       = TRUE
  ) |>
  flextable::footnote(
    i = 1,
    j = c(3, 5),
    ref_symbols = c("1", "2"),
    part = "header",
    value = flextable::as_paragraph(
      c(
        "The total occurrence of polypharmacy on the population level/perscription-day",
        "The expected occurrence of polypharmacy per person/perscription-day",
        "Prescription day is defined as the number of prescribed medication per day"
      )
    )
  ) |>
  flextable::align(align = "center", part = "all") |>
  flextable::align(j = 1:2, align = "left", part = "all") |>
  flextable::fontsize(size = 7, part = "all") |>
  #flextable::font(fontname = "Times", part = "all") |>
  flextable::width(j = c(1:8), c(1, rep(1.8, 7)), unit = "cm") |>
  flextable::set_table_properties(opts_pdf = list("arraystretch" = 1, "tabcolsep" = 0))

The highest average of prescription claim occurred on `r hi_claim$day`, with an average of `r hi_mean` [SD: `r hi_claim$sd_claim`] prescriptions. Meanwhile, the lowest average of claim occurred on `r lo_claim$day`, with an average of `r lo_mean` [SD: `r lo_claim$sd_claim`] prescriptions. A large margin of difference between the highest and lowest average number of claim implies daily variation and cyclicality, where it peaks on Monday, starts declining afterwards, and is at the lowest during weekends, as shown in table [-@tbl-overview-daily] and figure [-@fig-period-day]. The cyclical patterns are further explained in our supplementary material. @tbl-desc shows the number of prescriptions for each medication class alongside the distribution of DDD and its weight.

```{r tbl-overview-daily}
#| tbl-cap: "Daily prescription from 2018 to 2022 among patients with at least one claim of antidepressants or anxiolytics"

res_tbl_overview$overview |>
  dplyr::mutate(
    "mean_sd" = sprintf("%s (%.2f)", prettyNum(mean_claim, big.mark = ","), sd_claim),
    "median_IQR" = sprintf("%s (%.2f)", prettyNum(median_claim, big.mark = ","), IQR_claim),
    "Range" = sprintf("%s - %s", prettyNum(min_claim, big.mark = ","), prettyNum(max_claim, big.mark = ","))
  ) |>
  dplyr::select(day, mean_sd, median_IQR, Range) |>
  flextable::flextable() |>
  flextable::set_header_labels(
    day = "Day",
    mean_sd = "Mean (SD)",
    median_IQR = "Median (IQR)"
  ) |>
  flextable::add_header_row(
    values    = c("", "Number of Claims"),
    colwidths = c(1, 3),
    top       = TRUE
  ) |>
  flextable::align(align = "center", part = "all") |>
  flextable::align(j = 1, align = "left", part = "all") |>
  #flextable::font(fontname = "Times", part = "all") |>
  flextable::fontsize(size = 7, part = "all") |>
  flextable::autofit(add_w = 0, add_h = 0) |>
  flextable::set_table_properties(opts_pdf = list("arraystretch" = 1, "tabcolsep" = 0))

## Time-series decomposition

Using the weekly-aggregated data, we obtained decomposed series using an SSA-based approach. @fig-ssa depicts the decomposition result of the number of claim time-series in antidepressants. SSA-based decomposition separates the trend from its harmonics, as shown in the lower half of the plot. The trend explains roughly 90% of the variability, and harmonics explain 10%. The first two harmonics, F1 and F2, captured most of the variability in a series compared to the rest of it.

```{r fig-ssa}
#| fig-width: 16
#| fig-height: 12
#| fig-pos: H
#| fig-cap: An example of time-series decomposition using singular spectrum analysis-based approach

tar_read(plt_ssa_recon_n_claim)[[17]]

## Relative importance score

tar_load(res_recon_stat)

antidepressants <- res_recon_stat %>% subset(.$group == "Antidepressants")
anxiolytics     <- res_recon_stat %>% subset(.$group == "Anxiolytics")

respiratory <- res_recon_stat %>% subset(.$group == "Respiratory")
analgesics  <- res_recon_stat %>% subset(.$group == "Analgesics")

reportMedClaim <- function(med_obj) {
  res <- with(med_obj, prettyNum(n_claim, big.mark = ","))

  return(res)
}

reportMedEigen <- function(med_obj) {
  res <- with(med_obj, sprintf("%.4f [SD: %.4f]", eigen_mean, eigen_sd))

  return(res)
}

diffMed <- function(med_obj1, med_obj2, FUN) {
  med_rep1 <- FUN(med_obj1)
  med_rep2 <- FUN(med_obj2)

  res <- sprintf("%s vs %s", med_rep1, med_rep2)

  return(res)
}

Seven medications had a high relative importance as suggested by figure [-@fig-hi-eigen] and table [-@tbl-desc]. It is notable that highly prescribed medications usually have a high eigenvector centrality as well, which is theoretically supported by equation [-@eq-eigen-centrality]. However, despite higher prescription numbers (`r diffMed(antidepressants, respiratory, reportMedClaim)`), antidepressants had lower eigenvector centrality than medications for the respiratory system (`r diffMed(antidepressants, respiratory, reportMedEigen)`). We confirmed similar findings with anxiolytics, which had higher prescription numbers (`r diffMed(anxiolytics, analgesics, reportMedClaim)`) but lower eigenvector centrality (`r diffMed(anxiolytics, analgesics, reportMedEigen)`) than analgesics. The difference of eigenvector centrality became more pronounced as the number of prescriptions decreased.

```{r tbl-desc}
#| tbl-cap: Descriptive statistics of prescription data from 2018 to 2022, ordered by Eigenvector centrality as an indicator of nodal importance

res_recon_stat |>
  dplyr::arrange(-eigen_mean) |>
  dplyr::mutate(
    "mean_sd_eigen"     = sprintf("%.3f [%.3f]", eigen_mean, eigen_sd),
    "median_IQR_eigen"  = sprintf("%.3f [%.3f]", eigen_median, eigen_IQR),
    "mean_sd_dose"      = sprintf("%.2f [%.2f]", dose_mean, dose_sd),
    "median_IQR_dose"   = sprintf("%.2f [%.2f]", dose_median, dose_IQR),
    "mean_sd_weight"    = sprintf("%.2f [%.2f]", weight_mean, weight_sd),
    "median_IQR_weight" = sprintf("%.2f [%.2f]", weight_median, weight_IQR),
    "range_dose"        = sprintf("%.2f - %.2f", dose_min, dose_max),
    "range_weight"      = sprintf("%.2f - %.2f", weight_min, weight_max)
  ) |>
  subset(select = c(group, n_claim, mean_sd_eigen, mean_sd_dose, median_IQR_dose, mean_sd_weight)) |>
  flextable::flextable() |>
  flextable::set_header_labels(
    group            = "",
    n_claim          = "Claim",
    mean_sd_eigen    = "Mean [SD]",
    mean_sd_dose     = "Mean [SD]",
    median_IQR_dose  = "Median [IQR]",
    mean_sd_weight   = "Mean [SD]"
  ) |>
  flextable::add_header_row(
    values    = c("", "Importance", "Defined Daily Dose", "Weight"),
    colwidths = c(2, 1, 2, 1),
    top       = TRUE
  ) |>
  flextable::align(align = "center", part = "all") |>
  flextable::align(j = 1, align = "left", part = "body") |>
  #flextable::font(fontname = "Times", part = "all") |>
  flextable::fontsize(size = 7, part = "all") |>
  flextable::autofit(add_w = 0, add_h = 0) |>
  flextable::set_table_properties(opts_pdf = list("arraystretch" = 1, "tabcolsep" = 0))

```{r fig-hi-eigen}
#| fig-cap: Clusters of medication with a higher eigenvector centrality, implying a higher relative importance in polypharmacy regimens
#| fig-width: 9
#| fig-height: 10

tar_read(plt_hi_eigen_box)

# Discussion

@Delara2022

- The prevalence of polypharmacy in all medication classes is 37% [95% CI: 31-43%]
- Difference of prevalence measure is due to different numerical thresholds, study setting, and publication year

@Sirois2016

- Problem of exposure: Polypharmacy can be defined as cumulative or simultaneous prescriptions
- Cut-off threshold of polypharmacy should be justified by the risk of deleterious health outcomes
- An ideal monitoring system should describe the importance of polypharmacy

@Menditto2019

- Although polypharmacy is mostly researched in geriatric patients, younger population are also at risk of inappropriate multiple medication uses

@Shrivastava2013

- Among diagnoses, polypharmacy is often found in patients with schizophrenia, schizotypal, and delusional disorder
- Multi-class polypharmacy is the most common (20.9%), usually the combination of SSRI and benzodiazepine or TCA and benzodiazepine
- Five factors associated with psychiatric polypharmacy:
  - Scientific: Emphasis on biogenic amines roles in depression and schizophrenia expanded the early clinical evidence of psychotropic medications
  - Clinical: Standardization attempt resulted in overlapping diagnoses with limited validity; poly-nosology an polypharmacy go hand-in-hand
  - Economic: Pharmaceutical company influenced the demand of psychopharmaca uses
  - Political: US FDA imposes minimal guidelines for drug approval only for specific conditions
  - Cultural: Large cultural appetite for pharmacological treatments

@Stassen2021

- Attending psychiatrist is the main determinant of polypharmaca prescription
- Previous medical history and case severity at baseline only play a minor role in polypharmaca
- Adverse drug reaction in polypharmaca is 2-3 times higher compared to monotherapy, and not attributed to particular drug or drug combination; it implies that any drug and its combination can virtually trigger the side effect
- No advantage of polypharmacy over monotherapy, suggesting that polypharmacy regimens are not equally suited for every patient

@Molokhia2017

- The alternative measure is pre-defined criteria, e.g. BEERS, STOPP, and START
- Each additional medication contributes to 16% increased risk of errors

@Cavallo2012

- The weight is the number of time a pair of medications is prescribed together
- The co-prescription network has a scale-free property, suggesting that the probability new links being formed is growing alongside the degree of the target node
- The scale-free property of the network might be explained by the physician's prescribing habit: The physician tends to prescribe medications with known medications

@Bazzoni2015

- Drug-prescription network is highly dense and clustered at the global level
- Drug-prescription network is organized into anatomically homogeneous modules at the local level

# References

>>>>

@@

```{r fig-period-day}
#| fig-width: 14
#| fig-height: 5
#| fig-cap: Daily cyclical pattern captured on a de-trended data

tar_read(plt_period_day)[[1]]

@@
